Reaction-Advection-Diffusion Models of Air Pollution

### 2024-04-28 paper.tex
Need to add sources

\documentclass{article}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{physics}
\usepackage{tikz}
\title{Reaction-Advection-Diffusion Models of Air Pollution}
\author{Daniel Henricks}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{code.png}
    \caption{A few of the configuration options available. The total $\text{PM}_{2.5}$ concentration is composed of multiple
    sources. The total deaths per ethnicity are also shown.}
    \label{fig:grid2}
 \end{figure}

@@

To obtain our results, we utilized an API created by the company WattTime. WattTime combines marginal emission rates with InMAP's results
to produce a figure representing the health damage per MWh. To better understand how marginal emissions rates impact damage to human health caused
by emissions, consider the states of California and Mississippi. California has a much cleaner grid with a higher access to renewable energy (see below), 
so if a spike in demand occurs for energy production, it is much easier to compensate that demand with renewable energy. However,
in Mississippi, only 3 percent of the energy produced is renewable. In addition, the state consumes three times more energy than it produces.

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{renewables.png}
    \caption{A snapshot of California's grid makeup according to CAISO, California's Independent Service Operator.}
    \label{fig:grid2}
 \end{figure}

>>>>

\documentclass{article}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{physics}
\usepackage{booktabs}
\usepackage{tikz}
\title{Reaction-Advection-Diffusion Models of Air Pollution}
\author{Daniel Henricks}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{code.png}
    \caption{A few of the configuration options available. The total $\text{PM}_{2.5}$ concentration is composed of multiple
    sources. The total deaths per ethnicity are also shown.}
    \label{fig:grid3}
 \end{figure}

@@

To obtain our results, we utilized an API created by the company WattTime. WattTime combines marginal emission rates with InMAP's results
to produce a figure representing the health damage per MWh (using the actuarial values for a human life). To better understand how marginal emissions rates impact damage to human health caused
by emissions, consider the states of California and Mississippi. California has a much cleaner grid with a higher access to renewable energy, 
so if a spike in demand occurs for energy production, it is much easier to compensate that demand with renewable energy. However,
in Mississippi, only 3 percent of the energy produced is renewable. In addition, the state consumes three times more energy than it produces.
These factors make the marginal emissions rates in Mississippi much higher than ones in California.

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{renewables.png}
    \caption{A snapshot of California's grid makeup according to CAISO, California's Independent System Operator.}
    \label{fig:grid4}
\end{figure}

\subsection{Pricing of energy and marginal operating emissions rates (MOERs).}

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\textwidth]{prices.png}
    \caption{Consider a simple market with one coal generator and one gas generator. Initially, 
    the cost of carbon is low, and bids from coal generators are cheaper than those from gas generators. 
    As emissions become more costly, the cost of producing coal becomes more expensive than the cost of producing gas. 
    The result of the higher allowance price is lower emissions, as intended. (Howison, 2012)}
    \label{fig:grid4}
\end{figure}

For energy to be a traded commodity, electricity markets have to be deregulated. Some of the first markets to do
this were Alberta, Canada, which was deregulated in January of 1996, and California, which was deregulated in April 1998 (Barlow, 2002).
One of the strange features of power markets is the fact that zero or negative energy prices can occur and that the variance
of prices is much higher than other markets. In California, power usage peaks in the summer, whereas in Alberta, power usage and prices peak in the
winter (due to the cold). Since the energy markets are still relatively new, many papers rely on the early assumption that energy prices
follow a diffusion model. One of the key factors of this model is the fact that the price of energy can spike at times when demand is much larger than
supply.

Historically, in the absence of the trading of renewable energy credits (RECs), more pollution-intensive fuels are the first to be
used when an increase in power is demanded (Howison, 2012), meaning that the marginal emissions rate is higher when there is no incentives to use
renewable energy. Therefore, in areas where the cost of producing an additional unit of nonrenewable energy is much smaller than the 
cost of producing renewable energy, the average marginal emissions rate will be higher. This explains why there is such a large difference
in the MOER between regions of the United States.

To better visualize the emissions in the United States, WattTime provides a GeoJSON file of all of the Balancing Authorities
in the United States. In each of these regions, a marginal emissions rate is calculated every five minutes. Marginal emissions rates
can range from about 600 pounds of carbon produced per megawatt-hour in less carbon intense regions to over 2,000 pounds of carbon 
in more carbon intense grids. Using Leaflet (a JavaScript library used to visualize maps), the following map was created:

\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{moer-map.png}
    \caption{A map of all the Balancing Authorities (BAs) in the United States. Areas with higher marginal emissions rates (colored in brown/dark brown)
    will have higher health damages due to air pollution due to a higher reliance on nonrenewable sources. Conversely, areas with a lower MOER (colored in green/light yellow)
    will have lower health damages due to air pollution.}
    \label{fig:grid5}
\end{figure}

Now, we have to combine this marginal emissions rate with the InMAP model. Luckily, WattTime already has an API endpoint
that does this. Using the same methods as before, we can create a map of selected Balancing Authorities and their respective health
damages per MWh.

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{health-map.png}
    \caption{A map of selected health damages in the United States. Red colors mean higher health damage, whereas green colors correspond to
    lower health damage. As of April 28th, 2024, at 4:30 PM CDT, the TVA region had a damage to human health of \$45.70 per MWh of energy produced versus
    a damage to human health of \$1.00 in California per MWh.}
    \label{fig:grid5}
\end{figure}

\subsection{Final results and statistics}

WattTime also provides historical health damages for most locations in the United States since 2022. Using a running average from January 1st, 2024, to April 1st, 2024,
we created a Python script to compute the average health damage per region.

\begin{table}[h]
    \centering
    \caption{Average Health Damage and MOER by Region for January 1 to April 1, 2024}
    \begin{tabular}{@{}lcc@{}}
    \toprule
    Region                                   & Health Damage  & MOER  \\ \midrule
    TVA (Tennessee, Mississippi)             & \$54.00                    & 1,171                     \\
    MISO Lower MS River (Louisiana, Arkansas) & \$60.60                    & 1,389                     \\
    CAISO NORTH (California)                 & \$10.50                    & 718                       \\ \bottomrule
    \end{tabular}
\end{table}

Some interesting statistics we can get from this are:

\begin{enumerate}
    \item The health damages in areas with higher poverty rates tend to be greater than in those with more wealth.
    \item Solar farms typically have a lifespan of 35-40 years. Assuming a lifespan of 40 years with no downtime due to operational errors,
    building 1 MW of solar energy in the TVA region will produce approximately 80,000 MWh of energy (see https://clearloop.us/tn-paris/). With an average health damage of \$54.00,
    the social benefit of building this solar farm will be approximately \$4,320,000.
    \item The average health damage in the selected Southeastern regions is over five times greater than that in California.
  \end{enumerate}

\section{Conclusion}

In this paper, we have explained how air pollution can be one of the most detrimental factors to human health. Through the verification
of various ordinary differential equation models, in addition to diffusion and advection models for air pollution, the foundations of InMAP
were explained. In addition, recent experimental data on air pollution was utilized to prove many established ideas regarding air pollution in the United States.
Due to a lack of relevant data, it remains to be seen how health damage attributed to air pollution worldwide compares to that in the United States.
